---
title: "02_plotting_consensus_peaks"
author: "Dakota Hunt"
date: "4/25/23"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
library(ggplot2)
source("../../util/my_class_functions.R")
source("../../util/_setup.R")

```

# loading in peak_features env objects from 01_create_consensus_peaks
```{r laod env objects}

load("results/peak_features.RData", verbose = T)

```


# Peaks per dbp
```{r plotting peak features}

 ggplot(num_peaks_df, aes(x = num_peaks)) + 
  geom_histogram(bins = 100)

# Save figure
ggsave("figures/num_peaks_hist.pdf")

```


# Plot genome coverage (total peak length) against the number of peaks for each DBP
```{r peaks vs coverage}

ggplot(num_peaks_df, aes(x = num_peaks, y = total_peak_length)) +
  geom_point() + 
  ylab("BP covered") +
  xlab("Number of peaks") +
  ggtitle("Peak count vs. Genome Coverage")

# Save figure
ggsave("figures/peak_num_vs_coverage.pdf")


```
Result: there is an approximately linear relationship between the number of peaks and total coverage, which is to be expected.

# What is the distribution of promoter overlaps versus gene-bodies (hint hist)
```{r peaks vs coverage}
 ggplot(num_peaks_df, aes(x = peaks_overlapping_promoters)) + 
  geom_histogram(bins =25)

 ggplot(num_peaks_df, aes(x = peaks_overlapping_genebody)) + 
  geom_histogram(bins =25)

                 
```


# plotting num peaks on promoters vs number of peaks total for each DBP

```{r number of DBPS on promoters}
# Plotting x as num_peaks and y  as the number overlapping promoters.

ggplot(num_peaks_df,
       aes(x = num_peaks, y = peaks_overlapping_promoters)) +
  geom_point() +
  
  geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se=F, formula = 'y ~ x',
              color = "#a8404c") +
  ylim(0,60100) +
  xlim(0,60100) +

  # adding labels
xlab("Peaks per DBP") +
  ylab("Number of peaks overlapping promoters") +
  ggtitle("Relationship Between Number of DBP Peaks and Promoter Overlaps")

ggsave("figures/peak_num_vs_promoter_coverage.pdf")

```
Result: saturation of binding events -- as you get more peaks 
you stop increasing binding to promoters -- probably saturated.

# plotting num peaks over gene bodies vs number of peaks total for each DBP
```{r peak coverage on gene bodies}

ggplot(num_peaks_df,
       aes(x = num_peaks, y = peaks_overlapping_genebody)) +
  xlab("Peaks per DBP") +
  ylab("Number of peaks overlapping genes") +
  ggtitle("Relationship Between Number of DBP Peaks and Gene Body Overlaps")+
  geom_point() +
  geom_abline(slope = 1, linetype="dashed") +
  geom_smooth(method = "lm", se=F, formula = 'y ~ x',
              color = "#a8404c") +
  ylim(0,60100) +
  xlim(0,60100)

# saving
ggsave("figures/peak_num_vs_gene_body_coverage.pdf")
```
Result: Gene bodies make up almost all the places of binding in the genome




# Density plot of binding events
Making a density plot of num DBPs bound per promoter

```{r density plot of DBP localization events}

ggplot(peak_occurence_df, aes(x = number_of_dbp)) +
geom_density(alpha = 0.2, color = "#424242", fill = "#424242") +
  xlab(expression("Number of DBPs")) +
  ylab(expression("Density")) +
  ggtitle("Promoter binding events",
          subtitle = "mRNA and lncRNA genes") 

# saving
ggsave("figures/num_binding_events_per_promoter.pdf")

```
Result: promoter binding is bimodal
Result: most promoters have upto 100 dbps but some super-binders have ~200+ dbs
# RESULT: There are two types of promoter binding - (1) normal (2) super-binders


# promoters with out binding events
Finding promoters that don't have any DBPs bound
```{r prmoters with out binding events}

unbound_promoters <- peak_occurence_df %>% 
  filter(peak_occurence_df$number_of_dbp < 1)

# how many are there?
nrow(unbound_promoters)
# so there are 9448 promoters that don't have binding evetns (~10%)

#  let's put it in a folder called results. We will always use this folder structure
write_csv(unbound_promoters, "results/unbound_promoters.csv")

```

# lncRNA versus mRNA promoter binding
Let's compare the binding patterns of lncRNA vs mRNA promoters.
```{r lncrna vs mrna promoter binding}

ggplot(num_peaks_df, aes(x = num_peaks)) +
  geom_point(aes(y = peaks_overlapping_lncrna_promoters), color = "red") +
  geom_point(aes(y = peaks_overlapping_mrna_promoters), color = "black") +
  geom_smooth(aes(y = peaks_overlapping_lncrna_promoters), method = "lm", se = FALSE, formula = "y ~ x") +
  geom_smooth(aes(y = peaks_overlapping_mrna_promoters), method = "lm", se = FALSE, formula = "y ~ x")

# saving
ggsave("figures/lncRNA-mRNA-promoter_binding.pdf")




```

# How many of these proteins are TFs? What is the most represented type of DBD?
```{r prmoters with out binding events}
#How many proteins are TFs?
peak_TF_table <- table(num_peaks_df$tf, useNA = "always")
peak_TF_table

#What is the most represented type of DBD?
peak_DBD_table <- table(num_peaks_df$dbd, useNA = "always")
peak_DBD_table


```
344 proteins are TFs, 63 are not, and 23 have a value of NA.
C2H2 zinc fingers are the most common type of DNA-binding domain

