---
title: "01_binding_vs_expression"
author: "JR"
date: "8/17/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
source("../../../../util/intersect_functions.R")
source("../../../../util/_setup.R")
library(ggpubr)
```

# load in chipseq data from analysis/01_peak_feautres

```{r loading in chipseq data}
load("../01_peak_features/results/peak_features.RData", verbose = T)
```

Next we want to load in our final_samplesheet from 01_DESEQ_counts.Rmd
# Reading in sample sheet

```{r read in sample_sheet}
# First let's read in the sample sheet to know what is what
samplesheet <- read_rds("../../../05_R_analyses/05_RNAseq/01_differential_expression/results/final_samplesheet.rds")
```

# reading in TPM values from Salmon for our analyses

```{r reading in salmon Salmon TPMs}

# reading in salmon tpm
salmon_tpm <- read.csv("../../../05_R_analyses/05_RNAseq/00_RNAseq_download_NF_core_pipeline/00_NF_CORE_RNAseq_Pipeline_run/results/salmon/salmon_merged_gene_tpm.csv")

# TPM table is in same order as samplesheet
tpm <- salmon_tpm %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_id", values_to = "tpm") %>%
  merge(samplesheet) %>%
  group_by(gene_id, condition) %>%
  summarize(tpm = mean(tpm, na.rm = T)) %>%
  pivot_wider(names_from = condition, values_from = tpm, names_prefix = "tpm_")
```



# Plotting binding versus expression

Now let's examine how binding effects expression.
We have published previously that the more binding
events at a promoter the more abundant the expression is.
Let's see if this holds for our current subset of DBPs

# reading in promoter peak features data.frame from 12_.Rmd
Now let's read in the promoter overlaps (peak_occurrence_df)
how many DBPs are bound at each promoter

```{r loading in peak features data frame }

# peak_occurence_df is loaded in our environment and contains how many DBPs bind to each promoter (rows)
peak_occurence_df <- read_csv("../01_peak_features/results/peak_occurence_dataframe.csv")
View(peak_occurence_df)

# Now we can merge in the TPM data to this data.frame
# This will merge on the on teh gene_id column
promoter_features_df <- merge(peak_occurence_df, tpm)

# saving this file
write.csv(promoter_features_df, "results/promoter_feature_df_tpm.csv")

# let's plot the number of DBPs bound as a densuty plot.
ggplot(promoter_features_df, aes(x = number_of_dbp)) +
  geom_density() 

# Let's save
  ggsave("figures/DBP_binding_density_plot.pdf")


```

```{r DBP promoter binding versus total RNA expression}

# plotting binding vs total RNA expression
ggplot(promoter_features_df, 
            aes(y = log2(tpm_homo_sapiens_hepg2 + 0.001), x = number_of_dbp, color = gene_type)) + 
geom_point(data = promoter_features_df %>% filter(tpm_homo_sapiens_hepg2 > 0.001),
             shape = 17, alpha = 0.7) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs")) +
  stat_cor() +
  geom_smooth(method = "lm") +
  
  
  # just making look pretty
  scale_x_continuous(expand = c(0,0)) +
  scale_color_manual(values = c("#a8404c", "#424242"), name = "Gene type") + 
  ggtitle("Expression vs. promoter binding events") + 
  xlab(expression('Number of TFs')) +
  ylab(expression(log[2](TPM))) 

ggsave("figures/binding_vs_expression_total_rna.pdf")
```
Result: 
(1) There is a linear trend with number of DBPS and expression levels
(2) There is a population of genes that have numerous DBPs with low expression


# Abundance of genes in each cellular fraction

```{r TPM of genes in each fraction}
# First we need to the tpm DF into a matrix

tpm_matrix <- tpm %>% 
  column_to_rownames("gene_id") %>%
  as.matrix()
tpm_scaled <- t(scale(t(tpm_matrix)))
tpm_scaled <- tpm_scaled[complete.cases(tpm_scaled),]


# WARNING TAKES 5 min
# plotting
new.env()
pdf("figures/heatmap_expression.pdf", height =49, width = 12)
pheatmap::pheatmap(tpm_scaled, show_rownames = FALSE)
graphics.off()

```
RESULT: 
(1) Most RNAs are abundant in the nucleus
(2) Some RNAs expressed in total that are not in other fractions


# Making a heatmap of nuclear versus cytoplasmic expression
```{r TPM of genes in nuc vs cyto fractions}
# Filter tpm_scaled
tpm_scaled_nuc_cyto <- tpm_scaled[,colnames(tpm_scaled) == "tpm_homo_sapiens_cytosolic_fraction"
                                  | colnames(tpm_scaled) == "tpm_homo_sapiens_nuclear_fraction"]
# plotting
new.env()
pdf("figures/nuc_cyto_heatmap_expression.pdf", height =49, width = 12)
pheatmap::pheatmap(tpm_scaled_nuc_cyto, show_rownames = FALSE)
graphics.off()

```


# Differential expression analysis

```{r read in deseq_results_df}
# First let's read in the sample sheet to know what is what
deseq_res_shrunken_df <- read_rds("../../../05_R_analyses/05_RNAseq/01_differential_expression/results/nuclear_cyto_results_df.rds")

nuclear_genes <- deseq_res_shrunken_df%>% 
  filter(log2FoldChange < -1, padj < 0.05) %>%
  as.data.frame()

# saving file
write_csv(nuclear_genes, "results/nuclear_genes.csv")

# Cytoplasmic genes:
cyto_genes <- deseq_res_shrunken_df %>%
  filter(log2FoldChange > 1, padj < 0.05)

# saving file
write_csv(cyto_genes, "results/cyto_genes.csv")

# How many total genes are significantly enriched in nucleus? 5978
nrow(nuclear_genes)

# Breakdown of lncRNA vs mRNA genes in nucleus
nuclear_lncRNA_genes <- nuclear_genes[nuclear_genes$gene_id %in% lncrna_gene_ids,]
nuclear_mRNA_genes <- nuclear_genes[nuclear_genes$gene_id %in% mrna_gene_ids,]
nuclear_other_genes <- nuclear_genes[!nuclear_genes$gene_id %in% mrna_gene_ids & !nuclear_genes$gene_id %in% lncrna_gene_ids, ]

# Saving as csv's
write_csv(nuclear_lncRNA_genes, "results/nuclear_lncRNA_genes.csv")
write_csv(nuclear_mRNA_genes, "results/nuclear_mRNA_genes.csv")
write_csv(nuclear_other_genes, "results/nuclear_other_genes.csv")

nrow(nuclear_lncRNA_genes)
# 2112
nrow(nuclear_mRNA_genes)
# 2558
nrow(nuclear_other_genes)
# 1308

#How many total genes are significantly enriched in cytosol? 5978
nrow(cyto_genes)

cyto_lncRNA_genes <- cyto_genes[cyto_genes$gene_id %in% lncrna_gene_ids,]
cyto_mRNA_genes <- cyto_genes[cyto_genes$gene_id %in% mrna_gene_ids,]
cyto_other_genes <- cyto_genes[!cyto_genes$gene_id %in% mrna_gene_ids & !cyto_genes$gene_id %in% lncrna_gene_ids, ]

# Saving as csv's
write_csv(cyto_lncRNA_genes, "results/cyto_lncRNA_genes.csv")
write_csv(cyto_mRNA_genes, "results/cyto_mRNA_genes.csv")
write_csv(cyto_other_genes, "results/cyto_other_genes.csv")


nrow(cyto_lncRNA_genes)
# 916
nrow(cyto_mRNA_genes)
# 3501
nrow(cyto_other_genes)
# 638


```


# Top enriched genes in nuclear and cytosolic fractions and gene ontology analysis
```{r}
#Sort dataframes and look at most different genes by log2FoldChange

# Top 10 nuclear mRNA genes
nuclear_mRNA_genes <- nuclear_mRNA_genes[order(nuclear_mRNA_genes$log2FoldChange, decreasing = FALSE),]
head(n = 10, nuclear_mRNA_genes)

# Top 10 nuclear lncRNA genes
nuclear_lncRNA_genes <- nuclear_lncRNA_genes[order(nuclear_lncRNA_genes$log2FoldChange, decreasing = FALSE),]
head(n = 10, nuclear_lncRNA_genes)

# Top 10 cyto mRNA genes (decreasing should be true because fold change is positive in this case)
cyto_mRNA_genes <- cyto_mRNA_genes[order(cyto_mRNA_genes$log2FoldChange, decreasing = TRUE),]
head(n = 10, cyto_mRNA_genes)

# Top 10 cyto lncRNA genes (decreasing should be true because fold change is positive in this case)
cyto_lncRNA_genes <- cyto_lncRNA_genes[order(cyto_lncRNA_genes$log2FoldChange, decreasing = TRUE),]
head(n = 10, cyto_lncRNA_genes)

```

Nuclear enriched mRNA genes are associated with ECM organization and collagen fibril organization GO Biological Processes, and ECM-receptor interaction is the most enriched KEGG pathway.

Cytosolic enriched mRNA genes are associated with translation pathways and the RNA-binding GO molecular function.


# Reservoirs
```{r}

# define reservoirs as genes with >200 dbps bound and <0.01 tpm
promoter_features_df$reservoir <- 
  as.numeric(promoter_features_df$number_of_dbp > 200 & 
               promoter_features_df$tpm_homo_sapiens_hepg2 < 0.01)

# Table $reservoir to see how many 
table(promoter_features_df$reservoir)

# So there are 199 genes with superbinding promoters and that are not expressed in HepG2 cells.

# Create df with reservoir genes and check there is 199 with nrow()
reservoir_df <- promoter_features_df %>% 
  filter(promoter_features_df$reservoir == 1)
nrow(reservoir_df)

# How many reservoir promoters are lncRNA vs mRNA?
table(reservoir_df$gene_type)

# Save reservoir dataframe
write_csv(reservoir_df, "results/reservoir_genes.csv")



```





























# Binding versus nuclear expression
Let's see if the binding versus expression holds in the nuclear fraction
```{r binding versus nuclear expression}

# Now let's make a similar plot for nuclear RNA abundance versus #DBPs bound to their promoter
ggplot(promoter_features_df, 
            aes(y = log2(tpm_homo_sapiens_nuclear_fraction + 0.001), x = number_of_dbp, color = gene_type)) + 
  geom_point(data = promoter_features_df %>% filter(tpm_homo_sapiens_nuclear_fraction > 0.001),
             shape = 17, alpha = 0.7) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs")) +
  stat_cor() +
  
  
  #making look nice
  scale_x_continuous(expand = c(0,0)) +
  scale_color_manual(values = c("#a8404c", "#424242"), name = "Gene type") + 
  ggtitle("Nuclear Expression vs. promoter binding events") + 
  xlab(expression('Number of TFs')) +
  ylab(expression(log[2](TPM))) 

  # saving figure
  ggsave("figures/nuclear_expression-vs-promoter_binding.pdf")
```
RESULT: looks very similar to total RNA binding versus expression


# Binding versus cytoplasmic expression

Next we will determine the DBP binding versus cytoplasmic expression
```{Binding versus cytoplasmic expression}

# Same thing just seeing if there is a difference of cyto RNAs versus DBPs on promoter
ggplot(promoter_features_df, 
            aes(y = log2(tpm_homo_sapiens_cytosolic_fraction + 0.001), x = number_of_dbp, color = gene_type)) + 
  geom_point(data = promoter_features_df %>% filter(tpm_homo_sapiens_cytosolic_fraction > 0.001),
             shape = 17, alpha = 0.7) +
  geom_smooth(method = 'gam', formula = y ~ s(x, bs = "cs")) +
  stat_cor() +
  
  
  # making look nice
  scale_x_continuous(expand = c(0,0)) +
  scale_color_manual(values = c("#a8404c", "#424242"), name = "Gene type") + 
  ggtitle("Cytoplasmic Expression vs. promoter binding events") + 
  xlab(expression('Number of TFs')) +
  ylab(expression(log[2](TPM))) 
  # saving figure
  ggsave("figures/cytoplasmic_expression-vs-promoter_binding.pdf")
  
```
RESULTS:
(1) everything seems to be low abundance 
(2) Some mRNAs are expressed in the nucleus -- we could look at this more later.
(3) The same linear trend holds but is driven by mostly low expression events.

# lncRNA versus mRNA expression in total RNA
Next we will directly test the lncRNA vs mRNA expression levels in total RNA. 

```{r determining lncRNA and mRNA expression levels in total RNA}

# plotting
ggplot(promoter_features_df, aes(x = log2(tpm_homo_sapiens_hepg2 + 0.01), color = gene_type))+
  geom_density()

# saving figure
ggsave("figures/mrna_lncrna_tpm_total_rna.pdf")

# let's also do the same for nuclear since lncRNAs are typically more nuclear
ggplot(promoter_features_df, aes(x = log2(tpm_homo_sapiens_nuclear_fraction + 0.01), color = gene_type))+
  geom_density()

# saving figure
ggsave("figures/mrna_lncrna_tpm_nuclear.pdf")
```
RESULT:
(1) This yet again confirms lncRNAs have lower expression levels than mRNAs.
(2) In the nuclear fraction it shift's to closer.
(3) lot's of mRNA with nuclear expression -- that seems odd



We have previously observed that k562 cells also exhibit high binding promoters
that are not expressed. We termed them 'reservoirs' as they are a reservoir
for many Dna-protein interaction sites. Based on the plot above we observed that
this phenomena also exists in hepG2 cells as well. 


RESULT:
(1) There are no promoters with 5 elements bound


Now that we have defined reservoirs in hepG2 cells, we next want to determine how many 
are similar genomic regions in k562 and hepG2.

```{r reading in K562 reservoirs}

# reading in k562 promoter_features_DF
k562_df <- read_csv("/scratch/Shares/rinnclass/CLASS_2023/data/data/2020_k562_promoter_peak_df.csv")

# next we want to merge the k562 adn Hepg2 DFs 
k562_df <- k562_df %>% 
  dplyr::select(gene_id, reservoir, conservative_reservoir, tpm, expression, tf_binding, promoter_mean_tpm, promoter_median_tpm, promoter_max_tpm) %>%
  dplyr::rename(k562_reservoir = reservoir, 
                k562_conservative_reservoir = conservative_reservoir,
                k562_expression = expression,
                k562_tpm = tpm,
                k562_tf_binding = tf_binding,
                k562_promoter_mean_tpm =  promoter_mean_tpm,
                k562_promoter_median_tpm = promoter_median_tpm,
                k562_promoter_median_tpm = promoter_median_tpm,
                k562_promoter_max_tpm = promoter_max_tpm)

# save this file in new format
write_csv(k562_df,"results/k562_df.csv")

# renaming promoter_features_df to hepg2_df
hepg2_df <- promoter_features_df %>%
  dplyr::select(gene_id, gene_name, tpm_homo_sapiens_hepg2, tpm_homo_sapiens_cytosolic_fraction, tpm_homo_sapiens_nuclear_fraction, tpm_homo_sapiens_insoluble_cytoplasmic_fraction, tpm_homo_sapiens_membrane_fraction, number_of_dbp, hepg2_reservoir) %>%
   
  dplyr::rename( tpm_total = tpm_homo_sapiens_hepg2,
                 tpm_cytosolic_fraction =  tpm_homo_sapiens_cytosolic_fraction,
                 tpm_nuclear_fraction = tpm_homo_sapiens_nuclear_fraction ,
                 tpm_insoluble_cytoplasmic_fraction = tpm_homo_sapiens_insoluble_cytoplasmic_fraction ,
                 tpm_membrane_fraction = tpm_homo_sapiens_membrane_fraction)

# let's save this handy file
write_csv(hepg2_df,"results/hepg2_df.csv")
  
# Let's merge the k562 reservoirs in with HEPG2_df
# Merges on Gene_id
hepg2_k562_promoter_features_df <- merge(hepg2_df, k562_df)

# Now saving
write_csv(hepg2_k562_promoter_features_df, "results/hepg2_k562_promoter_features_df.csv")

# Make a table of reservoir status
res_status <- hepg2_k562_promoter_features_df %>% 
  group_by(hepg2_reservoir, k562_reservoir, k562_conservative_reservoir) %>%
  summarize(count = n())

# saving for future
write_csv2(res_status, "results/reservoir_overlap_stats.csv")
```
Result:
(1) There are 345 reservoirs in both K562 and HEPG2
(2) There are 80 Hepg2 reservoirs that overlap "conservative" k562 reservoirs



# Writting out files

```{r saving files for future use}
# We can now write these out for safekeeping / use in other analyses
write_csv(promoter_features_df, "results/promoter_features_df.csv")
write_csv(tpm, "results/mean_tpm_per_condition.csv")
write_csv(samplesheet, "results/samplesheet.csv")
```

