---
title: "03_clustering"
author: "Dakota Hunt"
date: "4/25/2023"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = F)
library(tidyverse)
library(GenomicRanges)
library(ggplot2)
library(pheatmap)
source("../../util/my_class_functions.R")
source("../../util/_setup.R")

```

# loading in peak_features env objects from 01_create_consensus_peaks
```{r laod env objects}

load("../01_consensus_peaks/results/peak_features.RData", verbose = T)

#let's look at the promoter peak occurence matrix (takes a min to load)
# read.table will make this a data.frame but will keep row names.
promoter_peak_occurence_matrix <- read.table("../01_consensus_peaks/results/promoter_peak_occurrence_matrix.tsv")

# Converting to a matrix format for correlation analysis
promoter_peak_occurence_matrix <- as.matrix(promoter_peak_occurence_matrix)
# Ok we see this is a very large matrix of 1 and 0 if a DBP is bound to a promoter.
# We can use these 36,000+ vectors of genes with 483 values of bound (1) or unbound (0)



```

# creating distance matrix & dendrogram
```{r distance matrix and dendrogram}

# creating distance matrix
peak_occurence_dist <- dist(promoter_peak_occurrence_matrix, method = "binary")

# clustering distance values
bin_hier <- hclust(peak_occurence_dist, method = "complete")

# This is a lot of ggplot -- on purpose
# please read through what is happening and parameter usage

 ggdendro::ggdendrogram(bin_hier, rotate = FALSE,  size = 3, 
                       theme_dendro = TRUE) +
   # 90 degree rotation to right
   coord_flip() +
   scale_y_continuous() +
   # adds label
   scale_x_continuous(position = "top") +
   # subsection to labels in order of clustering
   # ? seq_along
   scale_x_continuous(breaks = seq_along(bin_hier$labels[bin_hier$order]),
                      
                      # adding labels that are in the order 'column'
             labels = bin_hier$labels[bin_hier$order], position = "top",
             expand = c(0,0)) +
   theme(axis.text.x = element_text(angle = 90, hjust  = 1)) + 
   theme(axis.text.y = element_text(angle = 0,hjust = 1)) +
   scale_y_reverse(expand = c(0.01, 0)) +
   theme(
     plot.background = element_blank(),
     panel.grid.major = element_blank(),
   panel.grid.minor = element_blank(),
     panel.border = element_blank()
   )

# Nice let's save this using 'ggsave': since we used ggplot

ggsave("figures/ggdendro_plot.pdf", height = 50, width = 12, limitsize = F)


```
# Clustering of lncRNA and mRNA seperately
Now let's compare how lncRNAs and mRNAs cluster differnently?
Strategy: load annotations then index peak_occurrence for specific annotations

# lncRNA promoters ggdendro
```{r lncRNA promoter clustering}

lncrna_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_type %in% "lncRNA"] 

# Now we will use indexing to separate peak_occurrence_matrix in lncRNA and mRNA.
# note we are indexing and using indexing in indexing to get what we want.
lncrna_peak_occurence <- promoter_peak_occurence_matrix[,lncrna_promoters$gene_id]

# we do the clutering the same as above or we can combine the dist and hclust:
bin_hier_lncrna <- hclust(dist(lncrna_peak_occurence, method = "binary"))

# Now plot with ggdendro
ggdendro::ggdendrogram(bin_hier_lncrna, rotate = T,  size = 3)
 
# Now let's save this figure
ggsave("figures/lncrna_hclust_binary_dist.pdf", height = 49, width = 6)


```

# mRNA promter ggdendro
Now for mRNA

```{R mRNA promoter clustering}

mrna_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_type %in% "protein_coding"] 


# same strategy used very often:
# indexing into larger file and filtering to a file of stuff you want.
mrna_peak_occurence <- promoter_peak_occurence_matrix[,mrna_promoters$gene_id]

# getting the distance matrix for only mRNA promoters  
bin_hier_mrna <- hclust(dist(mrna_peak_occurence, method = "binary"))
 
# plotting with ggdendro
ggdendro::ggdendrogram(bin_hier, rotate = TRUE,  size = 3)

# saving
ggsave("figures/mrna_hclust_binary_dist.pdf", height = 44, width = 6)

# Cool we are making some good figures -- now let's dig into the biology
# Find an unknown protein cluster that has a known protein gene (hint ZNF)
# read about those genes and see if you could predict what the unknown gene is doing.

```
























# Using profile_tss for all 430 DBPs
# ! this takes ~45 min !
```{r metaplot DF of binding profiles over promoter}

# establishing DF
metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())

# for loop to populate DF 
for(i in 1:length(filtered_consensus_list)) {
  print(names(filtered_consensus_list)[[i]])
  tmp_df <- profile_tss(filtered_consensus_list[[i]], lncrna_mrna_promoters)
  tmp_df$dbp <- names(filtered_consensus_list)[[i]]
  metaplot_df <- bind_rows(metaplot_df, tmp_df)
  
}

# saving
write_rds(metaplot_df, "results/metaplot_df_final.rds")

```
  

# creating distance matrix of binding profile correlations
```{r scaling and plotting dendrogram of binding similarity by promoter}

metaplot_filtered_matrix <- metaplot_df %>% 
  pivot_wider(names_from = x, values_from = dens) %>%
  column_to_rownames("dbp") %>%
  as.matrix()
mm_scaled <- t(scale(t(metaplot_filtered_matrix)))
metaplot_hclust <- hclust(dist(mm_scaled), method = "complete")

# plotting relationship between binding profiles
plot(metaplot_hclust)
pdf("figures/tss_profile_dendrogram.pdf", height = 10, width = 27)
par(cex=0.3)
plot(metaplot_hclust)
dev.off()
```

# heat map of binding profile over +/- 1kb TSS window
```{r heatmap of profile over TSS (+/- 1Kb)}

# setting up PDF function for saving
pdf("figures/tss_profile_heatmap.pdf", height = 35, width = 10)

# complex heatmap of mm_scaled
Heatmap(mm_scaled, cluster_columns = FALSE, col = col_fun, border = TRUE, 
        show_column_names = FALSE,
        use_raster = TRUE,
        column_split = split,
        column_gap = unit(0, "mm"),row_names_gp = gpar(fontsize = 7))
# to save
dev.off()
```

# looking more into the patterns of ChIP binding profiles relative to TSS
# breaking this down into several clusters and plotting each
```{r how many groups are there?}


# Let's see how many DBPs have different patterns.
# manually checking at 6 we get a new group of 10 - after that not much changes
clusters <- cutree(metaplot_hclust, k=6)
table(clusters)

# plot cluter 1
dev.new()
pdf("figures/cluster_1_heatmap.pdf", height = 35, width = 10)

Heatmap(mm_scaled[clusters == 1,], cluster_columns = FALSE, col = col_fun, border = TRUE, 
        show_column_names = FALSE,
        use_raster = TRUE,
        column_split = split,
        column_gap = unit(0, "mm"),row_names_gp = gpar(fontsize = 7))
graphics.off()

# TSS enriched a bit broader than cluster 2


# cluster 2

dev.new()
pdf("figures/cluster_2_heatmap.pdf", height = 35, width = 10)

Heatmap(mm_scaled[clusters == 2,], cluster_columns = FALSE, col = col_fun, border = TRUE, 
        show_column_names = FALSE,
        use_raster = TRUE,
        column_split = split,
        column_gap = unit(0, "mm"),row_names_gp = gpar(fontsize = 7))
graphics.off()

# TSS enrichred a bit more narrow

# cluster 3
dev.new()
pdf("figures/cluster_3_heatmap.pdf", height = 35, width = 10)

Heatmap(mm_scaled[clusters == 3,], cluster_columns = FALSE, col = col_fun, border = TRUE, 
        show_column_names = FALSE,
        use_raster = TRUE,
        column_split = split,
        column_gap = unit(0, "mm"),row_names_gp = gpar(fontsize = 7))
graphics.off()

# very narrow binding over tss !

# cluster 4

dev.new()
pdf("figures/cluster_4_heatmap.pdf", height = 35, width = 10)
Heatmap(mm_scaled[clusters == 4,], cluster_columns = FALSE, col = col_fun, border = TRUE, 
        show_column_names = FALSE,
        use_raster = TRUE,
        column_split = split,
        column_gap = unit(0, "mm"),row_names_gp = gpar(fontsize = 7))
graphics.off()

# looks like TSS depletion and histone mods

# cluster 5 only 1 DBP
names(clusters[5])


# cluster 6
dev.new()
pdf("figures/cluster_6_heatmap.pdf", height = 35, width = 10)
Heatmap(mm_scaled[clusters == 6,], cluster_columns = FALSE, col = col_fun, border = TRUE, 
        show_column_names = FALSE,
        use_raster = TRUE,
        column_split = split,
        column_gap = unit(0, "mm"),row_names_gp = gpar(fontsize = 7))
graphics.off()

# TSS depletion and histone mods again.


```

# establishing lncRNA and mRNA promoters (+/- 1kb)
```{r create lncRNA and mRNA promoters }

# creating promoters just in case:
lncrna_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_type == "lncRNA"]
mrna_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_type == "protein_coding"]


```

# metaplots for each DBP by lncRNA and mRNA promoters
```{r}

#setting up lncrna DF.
lncrna_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())

# for loop to populate DF with overlap density in lncrna promoters
for(i in 1:length(filtered_consensus_list)) {
  print(names(filtered_consensus_list)[[i]])
  tmp_df <- profile_tss(filtered_consensus_list[[i]], lncrna_mrna_promoters = lncrna_promoters)
  tmp_df$dbp <- names(filtered_consensus_list)[[i]]
  lncrna_metaplot_df <- bind_rows(lncrna_metaplot_df, tmp_df)
  
}

# saving
write_rds(lncrna_metaplot_df, "results/lncRNA_metaplot_df_final.rds")

# now for mRNAs 
mrna_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())

# for loop to populate mRNA_metaplot
for(i in 1:length(filtered_consensus_list)) {
  print(names(filtered_consensus_list)[[i]])
  tmp_df <- profile_tss(filtered_consensus_list[[i]], lncrna_mrna_promoters = mrna_promoters)
  tmp_df$dbp <- names(filtered_consensus_list)[[i]]
  mrna_metaplot_df <- bind_rows(mrna_metaplot_df, tmp_df)
  
}

# saving mRNA metaplots
write_rds(mrna_metaplot_df, "results/mrna_metaplot_df_final.rds")


# now adding the information of gene type
mrna_metaplot_df$gene_type <- "mRNA"
lncrna_metaplot_df$gene_type <- "lncRNA"
combined_metaplot_profile <- bind_rows(mrna_metaplot_df, lncrna_metaplot_df)

# saving
write_rds(mrna_metaplot_df, "results/metaplot_df_final.rds")

#TODO why not printing
# TODO asign cluster value and plot the average of the clusters
# plotting - NOTE facet wrap by dbp !

# pdf(file = "figures/mega_plot_test.pdf")
ggplot(combined_metaplot_profile, 
       aes(x = x, y = dens, color = gene_type )) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  facet_wrap(dbp ~ ., scales = "free_y") +
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-1000, 0, 1000),
                     labels = c("-1kb", "TSS", "+1kb"),
                     name = "") + 
  ylab("Peak frequency") +
 scale_color_manual(values = c("#424242","#a8404c"))

# saving
ggsave("figures/mega_meta_plot_lncRNA-mRNA.pdf", width = 49, height = 12)
# ggsave("figures/mega_meta_plot_lncRNA-mRNA.pdf", limitsize = F)


```



# Make a metaplot of DBPS only on Super-binders versus regular promoters
```{r}
superbinder_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_id %in% superbinder_promoters_df$gene_id] 

normal_promoters <- lncrna_mrna_promoters[lncrna_mrna_promoters$gene_id %in% normal_promoters_df$gene_id] 

#setting up superbinder DF.
superbinder_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())

# for loop to populate DF with overlap density in lncrna promoters
for(i in 1:length(filtered_consensus_list)) {
  print(names(filtered_consensus_list)[[i]])
  tmp_df <- profile_tss(filtered_consensus_list[[i]], lncrna_mrna_promoters = superbinder_promoters)
  tmp_df$dbp <- names(filtered_consensus_list)[[i]]
  superbinder_metaplot_df <- bind_rows(superbinder_metaplot_df, tmp_df)
  
}

# saving
write_rds(superbinder_metaplot_df, "results/superbinder_metaplot_df_final.rds")

# now for normal promoters 
normal_metaplot_df <- data.frame(x = integer(), dens = numeric(), dbp = character())

# for loop to populate mRNA_metaplot
for(i in 1:length(filtered_consensus_list)) {
  print(names(filtered_consensus_list)[[i]])
  tmp_df <- profile_tss(filtered_consensus_list[[i]], lncrna_mrna_promoters = normal_promoters)
  tmp_df$dbp <- names(filtered_consensus_list)[[i]]
  normal_metaplot_df <- bind_rows(normal_metaplot_df, tmp_df)
  
}

# saving mRNA metaplots
write_rds(normal_metaplot_df, "results/normal_metaplot_df_final.rds")


# now adding the information of gene type
superbinder_metaplot_df$gene_type <- "superbinder"
normal_metaplot_df$gene_type <- "normal"
combined_super_normal_metaplot_profile <- bind_rows(superbinder_metaplot_df, normal_metaplot_df)

# saving
write_rds(combined_super_normal_metaplot_profile, "results/metaplot_df_final.rds")

#TODO why not printing
# TODO asign cluster value and plot the average of the clusters
# plotting - NOTE facet wrap by dbp !

# pdf(file = "figures/mega_plot_test.pdf")
ggplot(combined_metaplot_profile, 
       aes(x = x, y = dens, color = gene_type )) +
  geom_vline(xintercept = 0, lty = 2) + 
  geom_line(size = 1.5) + 
  facet_wrap(dbp ~ ., scales = "free_y") +
  ggtitle("Promoter Metaplot") + 
  scale_x_continuous(breaks = c(-1000, 0, 1000),
                     labels = c("-1kb", "TSS", "+1kb"),
                     name = "") + 
  ylab("Peak frequency") +
 scale_color_manual(values = c("#424242","#a8404c"))

# saving
ggsave("figures/mega_meta_plot_super-normal.pdf", width = 49, height = 12)
# ggsave("figures/mega_meta_plot_lncRNA-mRNA.pdf", limitsize = F)




```
